---
title: "COVID-19 in US counties"
author: "Alexandre Gouy"
date: "04/29/2020"
output: 
  # html_document
  rmdformats::readthedown:
    code_folding: hide
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

The first official case of **COVID-19** in the USA has been confirmed on the 21st of January. About three months later, almost 1 million cases have been discovered. In this context of this pandemic, it is of utmost importance to understand how the pandemic evolves by reporting data in a clear an insgihtful way.

This project has **two main goals**:

* **visualize different COVID-related metrics**: infection rate / 100k individuals, total cases, and total deaths at the county level (by representing maps and time series)

* **identify counties with potential errors in official counts**: it has been shown that some counties negative counts in cumulative cases, which is not possible.

Notes about the report: 

* R code used to generate this report is provided. You just need to click on the "Code" button on the right to display the code used in a given section.

* Most graphs and tables are interactive: you can zoom in and out, click on elements to display more content, or search for specific data points.

# Data preparation

## Dependencies and input files

We first need to load R packages (mostly used for interactive visualizations).

Then we set the location of input files used in this project:

* a web scraped data file (april15.csv) containing COVID data for a single time-point

* official population counts in US counties (census), directly taken from the web

* the New-York Times COVID-19 data, taken from the web

* official counties borders to be displayed on the map (downloaded from census.gov)

```{r dep, results = 'hide'}
## R packages to have installed
library(rgdal)
library(dplyr)
library(leaflet)
library(scales)
library(dygraphs)
library(crosstalk)
library(plotly)

## Local paths or URLs to files


```



